---
title: "Trachoma endgame serology: prepare individual-level dataset"
author: "Christine Tedijanto christine.tedijanto@ucsf.edu"
date: "Last updated: 03/01/22"
output:
    html_document:
      code_folding: hide
      toc: true
      toc_float: true
      toc_collapsed: true
      toc_depth: 3
number_sections: false
---

**Objective:** Clean data and create individual-level dataset for analysis.

```{r setup, message = FALSE}
library(here)
source(here("R/0-config.R"))

# save combined dataframe? (TRUE = save)
save_df <- TRUE

## Proctor studies -----
# WUHA
wuha_df <- read_csv(here(untouched_data_path, "wuha/20210923_wuha_ind_all.csv"))

# TAITU
taitu_df <- read_csv(here(untouched_data_path, "taitu/taitudata_christine.csv"))

# MORDOR
mordor_df <- read_rds(here(untouched_data_path, "mordor/R01-mordor-sero.rds"))

# PRET Niger
pret_df <- read_csv(here(untouched_data_path, "pret/Dry spot_PRET7 Niger Study Results-updated(9-15-2014).csv"))

## publicly available -----
# Kongwa (Odonkor 2021)
# downloaded from https://journals.plos.org/plosntds/article?id=10.1371/journal.pntd.0009343 supplement on 08/30/2021
kongwa_2018_df <- readxl::read_xlsx(here(untouched_data_path, "kongwa-2018/PLoSNTD_Odonkor_2021_Kongwa.xlsx"))

# Malawi (Burr 2019)
# downloaded from https://journals.plos.org/plosntds/article?id=10.1371/journal.pntd.0007749 supplement on 01/28/2022
malawi_df <- read_csv(here(untouched_data_path, "malawi/pntd.0007749.s001.csv"))

## other datasets -----
# Carter Center (Nash 2020)
cc_df <- read_csv(here(untouched_data_path, "carter-center/West Amhara serology data for SCR.TCC 2020.csv"))

# Kongwa (Wilson 2019)
kongwa_df <- read_rds(here(untouched_data_path, "kongwa/R01-kongwa-sero.rds"))
kongwa_pcr_df <- readxl::read_xlsx(here(untouched_data_path, "kongwa/Gates T1B 2012 Demographics and DBS and PCR.xlsx"), sheet = "PCR") # individual-level PCR for baseline
kongwa_pcr_clu1 <- readxl::read_xlsx(here(untouched_data_path, "kongwa/Copy of Baseline_2 year data with balozi.xlsx"), sheet = "Baseline") # cluster-level PCR at baseline
kongwa_pcr_clu2 <- readxl::read_xlsx(here(untouched_data_path, "kongwa/Copy of Baseline_2 year data with balozi.xlsx"), sheet = "2 year") # cluster-level PCR at year 2
kongwa_pcr_key <- read_rds(here(untouched_data_path, "kongwa/kongwa-mba-kids.rds")) %>% # link pcr ID to public ID
  mutate(id = row_number()) # to link to R01 dataset (see R01-trachoma-sero-kongwa.Rmd)
```

## Data cleaning

### WUHA

```{r wuha}
wuha_clean <- wuha_df %>% 
  filter(random_sample == 1, # include only children randomly sampled in study design
         age_group != "10+y",
         !(age_group == "6-9y" & survey %in% c(12,24)), # serology not collected for random sample at survey 12 and 24
         !(age_years>5 & survey %in% c(12,24)), # age years variable may not match with age_group variable in some cases (removes 94/7287, 1.3% rows)
         age_years <= 9, age_years > 0) %>%
  mutate(source = "WUHA",
         location_name = "Wag Hemra, Ethiopia (WUHA)",
         # based on ROC cutoff from reference samples
         # see Migchelsen et al. PLOS NTD 2017. https://doi.org/10.1371/journal.pntd.0005230
         pgp3_cutoff = 1113,
         ct694_cutoff = 337,
         # check years with Jeremy
         # based on Aragie et al. in Lancet Glob Health, study was from Nov 2015-Mar 2019 including census which occurred ~1 month before data collection
         year = case_when(survey == 0 ~ 2016,
                          survey == 12 ~ 2017,
                          survey == 24 ~ 2018,
                          survey == 36 ~ 2019),
         mda = case_when(survey == 0 ~ 1, TRUE ~ 0)) %>% 
  dplyr::select(source, location_name,
                cluster_id, household_id, individual_id,
                survey, year,
                age_months, age_years,
                pcr = pcr_individual, tf, ti,
                pgp3_cont = Pgp3, pgp3_cutoff, pgp3_bin = Pgp3_pos, 
                ct694_cont = Ct694, ct694_cutoff, ct694_bin = Ct694_pos,
                mda)

print(paste(nrow(wuha_clean), "rows in cleaned WUHA data")) #6687
```

### TAITU

```{r taitu}
taitu_clean <- taitu_df %>%
  filter(age.cc >= 12) %>% #remove children under age 1
  mutate(source = "TAITU",
         location_name = "Wag Hemra, Ethiopia (TAITU)",
         # based on ROC cutoff from reference samples
         # see Migchelsen et al. PLOS NTD 2017. https://doi.org/10.1371/journal.pntd.0005230
         pgp3_cutoff = 1113, 
         ct694_cutoff = 337,
         age_years = floor(age.cc / 12),
         year = 2018) %>% 
  # calculated binary for consistency w other studies
  # matches with "pos" variables in original TAITU dataset
  mutate(pgp3_bin = as.numeric(Pgp3>pgp3_cutoff),
         ct694_bin = as.numeric(Ct694>ct694_cutoff)) %>% 
  mutate(mda = case_when(
    studyarm %in% c("Targeted", "MDA") ~ 1,
    studyarm == "Delayed" ~ 0)) %>%
  mutate(pcr = case_when(
    pcr_ind == "Positive" ~ 1,
    pcr_ind == "Negative" ~ 0,
    pcr_ind == "Equivocal" ~ 0.5)) %>%
  # based on email communication with Jeremy - can assume all NA for clinical signs are 0
  mutate(tf = case_when(
    TF == "Yes" ~ 1,
    TRUE ~ 0)) %>% 
  mutate(ti = case_when(
    TI == "Yes" ~ 1,
    TRUE ~ 0)) %>% 
  dplyr::select(source, location_name,
                cluster_id = gwu, individual_id = masterperson,
                survey = phase, year,
                age_months = age.cc, age_years,
                pcr, tf, ti,
                pgp3_cont = Pgp3, pgp3_cutoff, pgp3_bin,
                ct694_cont = Ct694, ct694_cutoff, ct694_bin,
                mda)

print(paste(nrow(taitu_clean), "rows in cleaned TAITU data (month 24 only)")) #1487
```

### MORDOR

```{r mordor}
mordor_clean <- mordor_df %>% 
  filter(agem >= 12) %>% # remove children under 1 year old
  mutate(source = "MORDOR",
         temp_year = substr(testdate,1,4),
         age_years = floor(agem / 12),
         location_name = "Dosso, Niger") %>%
  # first 2 observations occurred 6 months apart
  mutate(year = case_when(
    phase == 0 ~ paste0("Spring ", temp_year),
    phase == 6 ~ paste0("Fall ", temp_year),
    TRUE ~ temp_year)) %>% 
  dplyr::select(source, location_name,
                cluster_id = clusterid, individual_id = childid,
                survey = phase, year,
                age_months = agem, age_years,
                antigen, cont = mfi, cutoff = serocut) %>%
  # data comes with one row per antigen, pivot wider to match other datasets
  pivot_wider(names_from = antigen,
              values_from = c(cont, cutoff),
              names_glue = "{antigen}_{.value}") %>% 
  mutate(pgp3_bin = as.numeric(pgp3_cont > pgp3_cutoff),
         ct694_bin = as.numeric(ct694_cont > ct694_cutoff),
         mda = NA) # mda administered for mortality every 6 months only to children 1-59 months

print(paste(nrow(mordor_clean), "rows in cleaned MORDOR data")) #4830
```

### PRET Niger

```{r pret}
pret_clean <- pret_df %>%
  filter(age > 0, age <= 5) %>% # one individual marked as age 6 removed
  dplyr::select(cluster_id = Code,
                individual_id = "Proctor ID",
                survey = VISIT,
                age_years = age,
                pgp3_cont = "pgp3 (22814) (49)",
                ct694_cont = "CT694 (712014) (46)") %>% 
  mutate(source = "PRET",
         year = 2013,
         # cutoffs based on ROC analysis
         # see Kim et al. PLOS NTD 2019
         pgp3_cutoff = 1083,
         ct694_cutoff = 496,
         location_name = "Matameye, Niger") %>% 
  mutate(pgp3_bin = as.numeric(pgp3_cont > pgp3_cutoff),
         ct694_bin = as.numeric(ct694_cont > ct694_cutoff)) %>% 
  mutate(individual_id = as.character(individual_id),
         mda = 1) # all clusters with serology received mda, either annual (6mo+) or biannual (<=12y)

print(paste(nrow(pret_clean), "rows in cleaned PRET data (serology at last time point only)")) #1010
```

### Kongwa (Odonkor 2021)

```{r kongwa_2018}
kongwa_2018_clean <- kongwa_2018_df %>% 
  mutate(source = "Kongwa (new)",
         location_name = "Kongwa, Tanzania",
         # cutoff based on ROC analysis using J-index
         # see Odonkor et al. PLOS NTD 2021
         pgp3_cutoff = 1771,
         year = 2018, 
         mda = 0) %>% # study was designed as survey 2 years post-MDA
  rename(tf_photo_right = "Photo Grade for TF, Right Eye (1=TF, 0=no TF)",
         tf_photo_left = "Photo Grade for TF, Left Eye (1=TF, 0=no TF)",
         tf_field_right = "Field Grade for TF, Right Eye (1=TF, 0=no TF)",
         tf_field_left = "Field Grade for TF, Left Eye (1=TF, 0=no TF)") %>% 
  # use field grades because many photo are NA
  # for field, one right and one left are marked with 9 -- assume this means missing and replace with 0
  mutate_at(vars(starts_with("tf_field")), ~ifelse(.==9, 0, .)) %>% # some values were >1, replace with 0
  rowwise() %>% 
  mutate(tf = as.numeric(max(tf_field_right, tf_field_left))) %>% 
  ungroup() %>% 
  dplyr::select(source, location_name,
                cluster_id = "Village ID",
                individual_id = "Participant ID",
                year, age_years = Age,
                pcr = "CT Infection (1=positive, 0=negative)",
                tf,
                pgp3_cont = "Mean fluorescence minus background (MFI-BG)",
                pgp3_cutoff,
                pgp3_bin = "pgp3 Seropositivity (1=yes, 0=no)",
                mda)

print(paste(nrow(kongwa_2018_clean), "rows in cleaned new Kongwa data")) #2401
```

### Malawi

```{r malawi}
malawi_clean <- malawi_df %>% 
  filter(EXAMINED==1, # yes with consent (see data dictionary, also in paper supplement)
         AGE<=9, AGE>=1) %>% 
  mutate(source = "Malawi", 
         year = 2014,
         district_name = case_when(District_binary == 0 ~ "Mchinji", District_binary == 1 ~ "Chikwawa"),
         mda = 1, # paper says that MDA was administered yearly from 2011-2013
         pcr = case_when(CT_PCR == 0 ~ 0, CT_PCR == 1 ~ 1), # 3 = invalid, 4 = missing
         tf_left = case_when(LEFT_EYE_TF == 0 ~ 0, LEFT_EYE_TF == 1 ~ 1), # 2 = not able to grade (only 1-2 of these)
         tf_right = case_when(RIGHT_EYE_TF == 0 ~ 0, RIGHT_EYE_TF == 1 ~ 1), # 2 = not able to grade (only 1-2 of these)
         ti_left = case_when(LEFT_EYE_TI == 0 ~ 0, LEFT_EYE_TI == 1 ~ 1), # 2 = not able to grade (only 1-2 of these)
         ti_right = case_when(RIGHT_EYE_TI == 0 ~ 0, RIGHT_EYE_TI == 1 ~ 1), # 2 = not able to grade (only 1-2 of these)
         cluster_id = as.character(CLUSTER)) %>% 
  rowwise() %>% 
  mutate(tf = max(tf_left, tf_right, na.rm = TRUE),
         ti = max(ti_left, ti_right, na.rm = TRUE),
         location_name = paste0(district_name, ", Malawi")) %>%
  ungroup() %>% 
  dplyr::select(source, location_name,
                cluster_id,
                household_id = INSTANCE_ID_HOUSE,
                individual_id = INSTANCE_ID_RES,
                year, age_years = AGE,
                pcr, tf, ti,
                pgp3_bin = ELISA_binary,
                pgp3_elisa = ELISA_A450, # elisa absorbance value
                mda)

print(paste(nrow(malawi_clean), "rows in cleaned Malawi data")) #6076
```

### Carter Center

```{r cc}
cc_clean <- cc_df %>% 
  mutate(source = "CarterCenter",
         year = 2017,
         cluster_id = as.factor(cluster),
         household_id = as.factor(hh_id),
         individual_id = as.factor(dbs_sample_id),
         location_name = paste0(district, ", Ethiopia"),
         # based on ROC analysis
         # see Nash et al. AJTMH 2020 (pg 3)
         pgp3_cutoff = 1558, 
         ct694_cutoff = 164) %>%
  mutate(mda = case_when(
    district == "Alefa" ~ 0, # had not received MDA for 2.5 years
    district %in% c("Andabet", "Dera", "Woreta town") ~ 1)) %>% # received MDA ~8 months prior
  dplyr::select(source, location_name,
                cluster_id, household_id, individual_id,
                year, age_years = age,
                pgp3_cont = pgp3_value, pgp3_cutoff, pgp3_bin = pgp3_binary, 
                ct694_cont = ct694_value, ct694_cutoff, ct694_bin = ct694_binary,
                mda)

print(paste(nrow(cc_clean), "rows in cleaned Carter Center data")) #2195
```

### Kongwa (Wilson 2019)

```{r kongwa}
## prepare pcr and TI data -----
kongwa_pcr_clean <- kongwa_pcr_df %>% 
  mutate(pcr = case_when(
    PCR_RESULT == "negative" ~ 0,
    PCR_RESULT == "positive" ~ 1)) %>% 
  dplyr::select(PATIENT_ID, pcr) %>% 
  mutate(svy = 1)

kongwa_addl <- kongwa_pcr_key %>% 
  left_join(kongwa_pcr_clean,
            by = c("seroid" = "PATIENT_ID", "svy")) %>% #seroid is internal ID, should be 902 rows with PCR
  mutate(tir = case_when(
    ti_r %in% c("1", "2", "Present", "Present (1 or 2)") ~ 1,
    ti_r %in% c("", "0", "0.5", "Absent", "Absent (0 or 0.5)", "NULL") ~ 0)) %>%
  mutate(til = case_when(
    ti_l %in% c("1", "2", "Present", "Present (1 or 2)") ~ 1,
    ti_l %in% c("", "0", "0.5", "Absent", "Absent (0 or 0.5)", "NULL") ~ 0)) %>% 
  mutate(ti = case_when(
    tir == 1 | til == 1 ~ 1,
    tir == 0 | til == 0 ~ 0)) %>% # let ti be NA if both values were missing
  dplyr::select(survey = svy, year, village_id = vilid,
                cluster_id = clusterid, household_id = hhid,
                mda = aztr, individual_id = id,
                ti, pcr) # also add ti

## clean and merge datasets -----
kongwa_clean <- kongwa_df %>%
  mutate(source = "Kongwa",
         cutoff = 10^roccut,
         cluster_id = clusterid,
         individual_id = id) %>% # note that this is a row ID, NOT a person ID
  mutate_at(vars(cluster_id, individual_id), as.character) %>% 
  dplyr::select(source, cluster_id,
                individual_id, survey = svy, year,
                age_years = age, 
                tf = tf, antigen,
                cont = mfi, # note that values in this dataset that were <= 0 were changed to 1
                cutoff) %>%
  # currently each antigen is its own row; pivot to make them columns
  pivot_wider(names_from = antigen,
              values_from = c(cont, cutoff),
              names_glue = "{antigen}_{.value}") %>% 
  mutate(pgp3_bin = as.numeric(pgp3_cont > pgp3_cutoff),
         ct694_bin = as.numeric(ct694_cont > ct694_cutoff),
         location_name = "Kongwa, Tanzania") %>% 
  left_join(kongwa_addl %>% mutate_at(vars(cluster_id, individual_id), as.character),
            by = c("survey", "year",
                   "cluster_id", "individual_id")) %>% 
  mutate_at(vars(tf,ti), as.numeric)

print(paste(nrow(kongwa_clean), "rows in cleaned Kongwa data")) #4989
```

## Combine and save data

```{r save}
## combine datasets
ind_df <- lapply(list(wuha_clean, taitu_clean, mordor_clean, pret_clean,
                      malawi_clean, kongwa_2018_clean,
                      cc_clean, kongwa_clean),
                 function(x) x %>% mutate_at(vars(year), as.character)) %>% 
  bind_rows() %>%
  # additional columns for all datasets
  mutate(location_year_name = paste(location_name, year)) %>% 
  mutate(location_year_name_clu = paste0(location_year_name, "_", cluster_id)) %>%
  # take log10 of continuous pgp3 measurement; if mfi-bg <= 0, set nonneg to 1 and log10 to 0
  mutate(sero_bin = pgp3_bin * ct694_bin) %>% # require both pgp3 and ct694 for sero pos
  mutate(pgp3_nonneg = ifelse(pgp3_cont<=0, 1, pgp3_cont)) %>% 
  mutate(pgp3_log10 = log10(pgp3_nonneg)) %>% 
  mutate(ct694_nonneg = ifelse(ct694_cont<=0, 1, ct694_cont)) %>% 
  mutate(ct694_log10 = log10(ct694_nonneg)) %>% 
  # reorder rows
  dplyr::select(source, village_id, cluster_id, household_id, individual_id,
                survey, year,
                location_name, location_year_name, location_year_name_clu,
                mda, age_months, age_years,
                pcr, tf, ti,
                pgp3_cont, pgp3_nonneg, pgp3_log10, pgp3_cutoff, pgp3_bin, pgp3_elisa,
                ct694_cont, ct694_nonneg, ct694_log10, ct694_cutoff, ct694_bin,
                sero_bin) %>% 
  # reorder lyn
  mutate(location_year_name = factor(location_year_name, levels = location_year_name_order))

print(paste0("Individual dataset has ", nrow(ind_df), " rows"))
head(ind_df)

## save
if(save_df){
  saveRDS(ind_df, file = here(temp_data_path, "ind_df.rds"))
}
```

## To resolve:

* Individual PCR results are not available for many 6-9 year olds in the first round of WUHA (published results based on pooled MLE). PCR results for 2016 will be skewed towards results for 0-5 year olds. 
* For TAITU, count all NA for TF/TI as 0? Changes TF from 66% overall to 54% and TI from 38% to 17%
* Add tests for comparing seroprevalence between arms for applicable studies
